A Guide to Designing Experiments to Test Statistical Graphics

### 2024-09-08 index.qmd
Updates to work with WIRE standards. Still to do: update refs with clickable DOIs, look for related wire articles, minimize lineup discussion in favor of other considerations, add introduction overview

---
title: A Guide to Designing Experiments to Test Statistical Graphics
author-list:
  - name: Emily Robinson
    dept: Statistics Department
    affiliation: California Polytechnic State University
    email: erobin17@calpoly.edu
  - name: Heike Hofmann
    dept: Statistics Department 
    affiliation: Iowa State University
    email: hofmann@iastate.edu
  - name: Susan Vanderplas
    dept: Statistics Department
    affiliation: University of Nebraska, Lincoln
    email: svanderplas2@unl.edu

abstract: |
  In this paper, we discuss considerations and methods for experimentally testing visualizations. We discuss levels of user engagement with graphics, common issues when developing a sampling or data generation model, the importance of pilot testing, and data analysis methods. Along the way, we also share hard-learned lessons on what not to do and some of the unique pitfalls of human testing in statistical and visualization research.

category: Research
bib: refs.bib
bibstyle: class-info/abbrv-doi.bst
# csl: ieee-transactions-on-visualization-and-computer-graphics.csl
ieee-keywords: 
  - \CCScatTwelve{Human-centered computing}{Visu\-al\-iza\-tion}{Visualization design and evaluation methods}{} 
  - \CCScatTwelve{Human-centered computing}{Visu\-al\-iza\-tion}{Empirical studies in visualization}{}
firstsection: Introduction

link-citations: false

format:  
  pdf:
    keep-tex: true
    template-partials: [partials/doc-class.tex, partials/title.tex, partials/before-body.tex, partials/after-body.tex]

pdf-engine: pdflatex

execute: 
  echo: false
  out-width: "\\columnwidth"

@@

date: last-modified
---

@@

Charts and data visualizations are primary methods of communicating scientific information to the public in what we typically hope is an easy-to-digest, visually attractive package.
There are many strategies for creating charts and graphs, from Tufte-esque minimalism\cite{tufte} to charts designed with extra imagery and aesthetic appeal that draw the viewer's attention and persist in memory\cite[Ch 3]{cairoFunctionalArtIntroduction2012}. 
For a specified type of data, there are also usually many different chart forms to display that data: for instance, if we have a set of categorical data and we wish to show the relative proportions of each category, we could do so using a stacked bar chart or the polar equivalent, a pie chart. 
There have been several attempts to list out all of the types of charts\cite{ribeccaSearchChartsData}, create a taxonomy of charts\cite{bertin1983semiology,desnoyersTaxonomyVisualsScience2011}, and even to classify charts using a domain-specific grammar of graphics\cite{wilkinsonGrammarGraphics1999}. 
With all of the different design choices available, how are chart creators to know what is the best approach for communicating data to the appropriate audience?

There are heuristics, rules-of-thumb, and various guidelines \cite{asme-standards-graphics,allenVisualizingScientificData2016,haemerDoubleScalesAre1948,fewInformationDashboardDesign2006,kosslynGraphDesignEye2006} for creating useful and visually attractive charts, but as scientists, we believe empirical evidence is essential for making correct decisions. 
The best way to establish the efficacy of various design decisions, then, is to test our charts by having them evaluated by people under controlled conditions and gathering appropriate data\cite{clevelandShapeParameterTwoVariable1988,clevelandGraphicalPerceptionGraphical1985}. 
It can be extremely challenging to design studies which strike the right combination of internal and external validity. 
Simply asking people to read quantities off of a graph may not generalize beyond the questions asked or the data used in the chart\cite{eellsRelativeMeritsCircles1926,croxtonBarChartsCircle1927,vonhuhnFurtherStudiesGraphic1927,croxtonGraphicComparisonsBars1932}, but designing a study that is sufficiently robust to those issues requires manipulation of so many factors as to be functionally close to impossible to complete.
In this paper, we attempt to distill lessons learned from several different sets of experimental graphics experiments\cite{clusters,sineillusion,emily-diss,power,vanderplasFramedReproducingRevisiting2019}, discussing the use of several different testing methods for statistical graphics\cite{vanderplasTestingStatisticalCharts2020}, the process of designing a graphical experiment, and analysis of the resulting empirical data. 
It is our hope that by writing this paper, we will prevent other researchers from repeating hard lessons in what to do (and what not to do), lowering the barriers that exist for conducting research involving empirical tests of statistical graphics.

@@

We will start with an overview of different testing methods that we have used to empirically assess graphics. 
While it is certainly possible to use equipment such as eye-tracking devices, here we will primarily focus on studies which can be conducted over the internet, because these experiments have the lowest overhead, offer relatively fast data collection, and provide useful results that can be very clear if the experiment is designed correctly.
Our toolkit is R-based\cite{R}, and includes ggplot2\cite{ggplot2} and Shiny\cite{shiny} as primary components, as well as  JavaScript (including D3\cite{d3}) to customize the Shiny interface for many of the experiments.

One primary concern when testing graphics is the level of user engagement required. 
For instance, if we are testing whether someone can detect an effect such as a trend line in noisy data, this is a *perceptual* question best addressed using methods which allow the user to interact with the data on a basic visual level. 
Sometimes, it may be possible to set up a scenario where the user adjusts the plot directly, as in our experiment testing the strength of the sine illusion experiment\cite{sineillusion}: users adjusted the strength of the illusion correction until the lines appeared to be the same length, providing a direct measure of the magnitude of the sine illusion's effect, as in @fig-sine-illusion.

@@

Statistical lineups\cite{bujaStatisticalInferenceExploratory2009,wickhamGraphicalInferenceInfovis2010} are a more generally applicable testing tool for perceptual questions such as "which chart displays this data more clearly"\cite{power} while simultaneously assessing the significance of the graphical finding in the chart. 
As in the criminal procedure, lineups embed a target plot with signal into an array of 19 other innocent "null" plots that show only noise; if viewers consistently pick the target plot at a higher rate than the 19 null plots, then the target plot is said to be visually significant\cite{majumderValidationVisualStatistical2013,vanderplasStatisticalSignificanceCalculations2021} and a "see" value (the visual analogue of a $p$-value for a statistical test) can be calculated\cite{niladriroychowdhurySeeValueApp2020}. 
<!-- Beyond the quantitative assessment of the signal with $p$- and see-values, the human involvement also allows us to collect information on the qualitative nature of *what* a viewer sees.--> 
In another variation of the statistical lineup procedure, two models are compared, with target plots from each model embedded in the array of 20 total plots; the remaining null plots are constructed from a mixture model that blends the two competing models. This variation allows the experimenter to assess graphical design choices to determine whether they effectively emphasize structural differences in the data and assign a hierarchy of visual attributes\cite{clusters}.
Statistical lineups do not require the experimenter to provide any contextual information for the data: all of the necessary information is embedded in the choice of a null statistical model, which is convenient for testing purposes, but does not require engagement at a level that would allow for testing the viewer's understanding of the information shown in the chart. 
An example of a statistical lineup for testing the perception of exponential growth is shown in @fig-lineup-log.

![A lineup experiment testing whether users can perceive differences in rates of exponential growth on linear and log scales. In addition to asking users to select the plot which is the most different, we also asked for basic reasoning and the user's level of confidence in their answer.](Lineup_log_screenshot.png){#fig-lineup-log}

To assess the viewer's *understanding* of information shown in a chart, we must ask questions and allow the user to provide feedback. User feedback may be collected on a numerical scale or through the use of written comments, recorded "think-aloud" processes, and other more qualitative interaction methods. 
We have conducted studies asking users to provide us with a written description of what the chart shows within a wider story-driven scenario (@fig-estimation-describe), and we have also asked for written feedback about why a certain panel of a statistical lineup was chosen; the insights from participant responses to these questions have been rich and insightful (and will be discussed more in the analysis section). 
While we have not (to date) recorded users talking out loud about what they are seeing, it is entirely feasible that so-called "think aloud" processes could be used within a Shiny application, with audio recordings saved to the server for transcription and analysis\cite{dunbar1995scientists,traftonTurningPicturesNumbers2000,kirschenbaum2003comparative}.

@@

Of course, as we have confined ourselves to the discussion of online, asynchronous testing methods, every user interaction with the testing materials (typically hosted on a web page) can also be recorded along with time stamps, mouse positions, browser size and screen resolution, and other information. While we have not used this type of information heavily in our experimental analyses thus far, we do collect some time stamp data in order to assess how long participants spend on each question. What we typically see, is that the first round of test questions takes the longest for participants to complete. Additional experiments do not affect accuracy (i.e. there is no immediate learning effect) until after `too many' tests cognitive fatigue proves to be detrimental to accuracy\cite{chowdhury2018}. This sweet spot between user replicates and fatigue depends on the cognitive burden in each test and should factor into designing the experiment.
<!--While we have not used this type of information heavily in our experimental analyses thus far, we do collect some time stamp data in order to assess how long participants spend on each question; it could almost certainly be more informative in the right setting.-->
In addition, we have also provided participants with supportive tools, such as "scratch pads" and limited function calculators built into the Shiny application that allow us to gain additional insight into the estimation and visual comparison process when more complex calculations are required to answer the numerical estimation question (@fig-estimation-calc). 
These supportive tools have provided a wealth of information that, while not easy to analyze and generally messy, provides real insight into how participants were looking at the data, what strategies they tried and discarded for reading the chart, and what visual estimation methods were used.

@@

One of the most difficult components of designing an experiment which asks users to directly estimate information from a chart using a full scenario (background information, etc. as well as contextual details from the chart) is that the questions must be extremely carefully constructed.
Mathematics education researchers provide guidelines for selecting different levels of questioning in order to assess graph comprehension: literal reading of the data, reading between the data, and reading beyond the data \cite{wood1968objectives, curcio1987comprehension, friel2001making, glazer2011challenges}. In a recent study, we identified questions based on this framework to evaluate direct estimates and extend those estimates to make comparisons between two points.

Even when great care is taken with the construction of the question, participant answer accuracy is fundamentally limited by the fact that many participants do not read and interpret the question with the care and precision that it was written.
Questions that ask participants to e.g. estimate the multiplicative change in a quantity at two time points may be misunderstood as asking for an estimate of the additive difference, and the resulting estimates are then one or more orders of magnitude off of the correct answer.
This is one area where lineup tests are convenient - they do not depend on participants to understand the nuances of language or scenarios built around the chart under investigation.
However, in some situations it may be sufficient to ask participants to estimate direct numerical quantities that have little contextual information, as we did in \cite{vanderplasFramedReproducingRevisiting2019} when assessing the accuracy of framed plots re-created from the Statistical Atlas.

It is also possible to design a task that requires participants to *engage directly* with an interactive visualization. 
This can be useful in a directed task, where users are asked to interact with the chart in a specific way and the result is recorded, but it is also possible to use interactive visualizations in an open-ended task, recording how users engage with the graphic in an exploratory (as opposed to goal-directed) manner. 
In one recent experiment, we asked participants to forecast an exponential trend, with data presented on either a linear or log scale. 
Using JavaScript code modified from New York Times interactive graphics "You Draw It" features\cite{katzYouDrawIt2017}, we had users draw trend lines with their computer mouse and make forecasts directly on interactive charts, with the data and user-drawn predictions recorded to our database.
With interactive graphics rendered using JavaScript (or other web libraries), the only limit to the types of questions one can ask in testing graphics is one's ability to write code to interact with the visualization library. 
This type of testing method can be extremely natural for participants, but it also is hard to generalize when discussing testing methods because of the potential range of applications where it might be employed.

@@

Once you have decided what graphical task you want to explore or test, it is necessary to develop a model which can be used to test the graphical features you want to explore in a precise manner. 
This is the single longest part of the entire experimental design and execution process, in part because choosing a model that replicates important visual features of the data is an extremely complex process \cite{hullmanDesigningInteractiveExploratory2021,cookFoundationAvailableThinking2021,vanderplasDesigningGraphicsRequires2021}.

There are two main options when developing a statistical model for graphical testing: you can either start with a large dataset and sample from that dataset\cite{power}, or you can start from a model and sample data from that model generating process\cite{sineillusion,clusters,emily-diss}.
This decision is largely determined by the availability of a large dataset containing the requisite features of interest and the qualities being manipulated in the experiment. For instance, Hofmann et al. \cite{power} used samples of different sizes from a pre-existing dataset to manipulate the amount of signal in each comparison; with a small sample, there is less signal and the same amount of noise, making the true plot harder to spot. 
In many situations, though, a convenient dataset with the right properties is harder to acquire, and it becomes necessary to develop a sampling model to generate data for user evaluation.

The tools we discuss in the remainder of this section can be applied both to pre-existing data sets and to model-based sampling methods. 
We will do our best to keep this discussion general, but we will provide details in some situations to ensure that our general descriptions are not hopelessly vague.

@@

Once an appropriate set of parameters are identified using the numerical screening method, then it is important to calibrate these parameter selections visually; after all, no numerical statistic is a perfect measure of what we actually see; at best, they are approximations of what we might potentially see. 
We have found it to be useful to have one experimenter calibrate the model parameters at a gross level, and then have another experimenter narrow in on the parameters which are visually reasonable within the selected range. Then, both examiners visually inspect a large number of plots generated using those parameters to get a sense for how difficult the task at hand is (this strategy is also described by Lu et al. \cite{luModelingJustNoticeable2022}). 
At some point, all experimenters become so visually saturated with the nuances of the data generating mechanism that it may become necessary to "sanity check" the protocol with family members, friends, and colleagues. 
These informal surveys provide extremely useful feedback and can help to counteract the visual saturation of being immersed in the design of a visualization experiment for months at a time.

@@

We cannot overstate the importance of visual assessment of your model stimuli, preferably with fresh eyes. We highly recommend performing a series of think-aloud pilot testing before deploying an experiment. In support of this assessment, we offer up a cautionary tale of our own experience: that of \cite{clusters}, where we designed an experiment to test which plot aesthetics promoted discovery of linear trends and/or clusters.

The experiment was a full 2x3x3 factorial exploration of three data generating parameters, with 3 replicates at each parameter combination (54 datasets) and 10 aesthetic combinations (for a total of 540 lineups). Each lineup had 20 different sub-panels, so we should have carefully visually inspected some 10800 different panels. 
As is evident from the fact that we're telling this story as a cautionary tale, we missed a critical problem with our data-generating mechanism: when clusters were assigned to randomly generated data after the fact, we didn't control the cluster size, leading to clusters of one or two points in relatively few sub-panels. 
This became particularly noticeable when bounding ellipses were added to the plot, as the method used to generate those ellipses required at least 3 points in the cluster. The missing boundary ellipse in the corresponding sub-panels escaped our notice during the stimuli proof-reading phase of the experiment, but did not escape the notice of our participants, who only needed to examine about 10 lineups each (around 200 panels). 
An example of one of the problematic lineups is shown in @fig-lineup-problems: many participants selected panel 16 because of the missing ellipse; not a wrong choice, but certainly not the effect we intended to test.

![A lineup from Vanderplas & Hofmann \cite{clusters}. Panel 10 shows the clustered target data and panel 17 shows the target data with a strong linear relationship; either of these target panels was the expected choice. Unfortunately, panel 16 has only two bounding ellipses shown, which is an unintentional difference that resulted from a faulty method for assigning clusters to null plots; many participants selected this panel instead of one of the target panels.](lineup-missing-ellipse.png){#fig-lineup-problems}

@@

While studies have found some relationship between lineup performance and demographic factors\cite{majumderHumanFactorsInfluencing2014, vanderplasSpatialReasoningData2016a}, these differences are relatively small and unlikely to be important. 
Studies have also not found a strong relationship between visual task performance and participant recruitment method\cite{vanderplasFramedReproducingRevisiting2019}, perhaps in part because demographics which use social media sites and demographics completing online research tasks for pay overlap heavily. 
What is clear, however, is that some participants are very good at evaluating graphics and some participants are not; as a result, it can be extremely helpful to use random effects models with an effect for individual participants.

@@

**Generalized Linear Mixed Models** 
 The see-value approach \cite{niladriroychowdhurySeeValueApp2020,vanderplasStatisticalSignificanceCalculations2021,majumderValidationVisualStatistical2013} is extremely useful for single lineups, but when a series of lineups that are part of a designed experiment are used, generalized linear mixed models are a much simpler way to summarize the overall effect of various manipulated factors\cite{power,clusters,emily-diss}. This approach also works extremely well for psychometric experiments \cite{sineillusion}, as psychometric models can be easily fit into the framework of a generalized linear mixed model. In addition, similar model structures can be used to model accuracy, response time, and confidence, if all three types of information are collected from participants during the experiment.

**Numerical Estimation** There are additional considerations that should be expected when asking participants to estimate numerical quantities. Anchoring and rounding cause participant responses to cluster in ways that can bias statistical estimators, requiring methods designed for these types of data\cite{ushakovStatisticalAnalysisRounded2017,tourangeau_rips_rasinski_2000,heitjanIgnorabilityCoarseData1991}. An alternative approach is to analyze the data graphically using e.g. a combination of density and rug geoms for point estimates (as shown in @fig-density-rug), or spaghetti plots for individual response curves, with annotations to show anchor points. The advantage of this approach is that it can accommodate extremely messy data without requiring the extensive data cleaning and elimination of nonsensical responses that might be necessary to fit a statistical model.

@@

grid_lines_data <- tibble(scale = c(rep("linear", 12), rep("log2", 10)), 
                          grid_lines = c(seq(0,55000, 5000),
                                         2^seq(7,16))
) %>%
  expand_grid(dataset = c("dataset1", "dataset2"))

@@

```{r density-plot, include = F, echo = F}
densityPlot <- function(data, datasetID, estimate, xlabel = "Estimate", x_limits = c(0,70000), zoom = F, scalesx = T, zoom_limits = c(NA,NA), gridlines = T, rugjitter = 0.25) {

  estPlot <- data %>%
    filter(dataset %in% datasetID) %>%
    ggplot(aes_string(x = estimate, fill = "scale", color = "scale")) +
    geom_density(alpha = 0.5, color = NA) +
    geom_rug(aes(y = -Inf), alpha = 0.6, show.legend = F, position = position_jitter(width = rugjitter, height = 0)) +
    geom_vline(aes(xintercept = true_value, linetype = "a")) +
    geom_vline(aes(xintercept = closest_pt_value, linetype = "b"))

  if (gridlines) {
  estPlot <- estPlot + 
    geom_vline(data = grid_lines_data %>% filter(dataset == "dataset1", grid_lines >= x_limits[1], grid_lines <= x_limits[2]),
               aes(xintercept = grid_lines, color = scale, linetype = "c"))
  }

>>>>

---
title: A Guide to Designing Experiments to Test Statistical Graphics
author:
  - name: Emily Robinson
    dept: Statistics Department
    affiliation: California Polytechnic State University
    email: erobin17@calpoly.edu
  - name: Heike Hofmann
    dept: Statistics Department 
    affiliation: Iowa State University
    email: hofmann@iastate.edu
  - name: Susan Vanderplas
    dept: Statistics Department
    affiliation: University of Nebraska, Lincoln
    email: svanderplas2@unl.edu

abstract: |
  In this paper, we discuss considerations and methods for experimentally testing visualizations. We discuss levels of user engagement with graphics, common issues when developing a sampling or data generation model, the importance of pilot testing, and data analysis methods. Along the way, we also provide recommendations of how to avoid some of the unique pitfalls of human testing in statistical and visualization research.

bibliography: refs.bib
csl: apa.csl

mainfont: Helvetica
monofont: Roboto
number-depth: 2
number-sections: true

execute: 
  echo: false

@@

date: last-modified

format:  
  pdf:
    keep-tex: true
    fontsize: 10pt
---

@@

# Introduction

Charts and data visualizations are primary methods of communicating scientific information to the public in what we typically hope is an easy-to-digest, visually attractive package.
There are many strategies for creating charts and graphs, from Tufte-esque minimalism [@tufte] to charts designed with extra imagery and aesthetic appeal that draw the viewer's attention and persist in memory [@cairoFunctionalArtIntroduction2012 Ch 3]. 
For a specified type of data, there are also usually many different chart forms to display that data: for instance, if we have a set of categorical data and we wish to show the relative proportions of each category, we could do so using a stacked bar chart or the polar equivalent, a pie chart. 
There have been several attempts to list out all of the types of charts [@ribeccaSearchChartsData], create a taxonomy of charts [@bertin1983semiology;@desnoyersTaxonomyVisualsScience2011], and even to classify charts using a domain-specific grammar of graphics [@wilkinsonGrammarGraphics1999]. 
With all of the different design choices available, how are chart creators to know what is the best approach for communicating data to the appropriate audience?

While there are heuristics, rules-of-thumb, and various guidelines [@asme-standards-graphics;@allenVisualizingScientificData2016;@haemerDoubleScalesAre1948;@fewInformationDashboardDesign2006;@kosslynGraphDesignEye2006] for creating useful and visually attractive charts, the best way to establish the efficacy of various design decisions is to test charts, evaluate them under controlled conditions, and gather appropriate data [@clevelandShapeParameterTwoVariable1988;@clevelandGraphicalPerceptionGraphical1985]. 
It can be extremely challenging to design studies which strike the right combination of internal and external validity. 
Simply asking people to read quantities off of a graph may not generalize beyond the questions asked or the data used in the chart [@eellsRelativeMeritsCircles1926;@croxtonBarChartsCircle1927;@vonhuhnFurtherStudiesGraphic1927;@croxtonGraphicComparisonsBars1932], but designing a study that is sufficiently robust to those issues requires manipulation of so many factors as to be functionally close to impossible to complete.
In this paper, we attempt to distill the experience gained from conducting several different types of graphics experiments [@clusters;@sineillusion;@emily-diss;@power;@vanderplasFramedReproducingRevisiting2019], discussing the use of different testing methods [@vanderplasTestingStatisticalCharts2020], the process of designing a graphical experiment, and analysis of the resulting empirical data. 
It is our hope that this paper will lower the barriers that exist for conducting empirical graphics research and reduce the probability of costly mistakes.

@@

There are many different testing methods used to empirically assess graphics. 
While it is certainly possible to use equipment such as eye-tracking devices, this paper focuses primarily focus on studies which can be conducted over the internet, because these experiments have the lowest overhead, offer relatively fast data collection, and provide useful results that can be very clear if the experiment is designed correctly.
The toolkit used here is R-based [@R], and includes ggplot2 [@ggplot2] and Shiny [@shiny] as primary components, as well as  JavaScript (including D3[@d3]) to customize the Shiny interface for many of the experiments.

One primary concern when testing graphics is the level of user engagement required. 
For instance, testing whether someone can detect an effect such as a trend line in noisy data is a *perceptual* question best addressed using methods which allow the user to interact with the data on a basic visual level. 
Sometimes, it may be possible to set up a scenario where the user adjusts the plot directly, as in an experiment testing the strength of the sine illusion experiment [@sineillusion]: users adjusted the strength of the illusion correction until the lines appeared to be the same length, providing a direct measure of the magnitude of the sine illusion's effect, as in @fig-sine-illusion.

@@

Statistical lineups [@bujaStatisticalInferenceExploratory2009;@wickhamGraphicalInferenceInfovis2010] are a more generally applicable testing tool for perceptual questions such as "which chart displays this data more clearly" [@power] while simultaneously assessing the significance of the graphical finding in the chart. 
As in the criminal procedure, lineups embed a target plot with signal into an array of 19 other innocent "null" plots that show only noise; if viewers consistently pick the target plot at a higher rate than the 19 null plots, then the target plot is said to be visually significant [@majumderValidationVisualStatistical2013;vanderplasStatisticalSignificanceCalculations2021] and a "see" value (the visual analogue of a $p$-value for a statistical test) can be calculated [@niladriroychowdhurySeeValueApp2020]. 
<!-- Beyond the quantitative assessment of the signal with $p$- and see-values, the human involvement also allows us to collect information on the qualitative nature of *what* a viewer sees.--> 
In another variation of the statistical lineup procedure, two models are compared, with target plots from each model embedded in the array of 20 total plots; the remaining null plots are constructed from a mixture model that blends the two competing models. This variation allows the experimenter to assess graphical design choices to determine whether they effectively emphasize structural differences in the data and assign a hierarchy of visual attributes [@clusters].
Statistical lineups do not require the experimenter to provide any contextual information for the data: all of the necessary information is embedded in the choice of a null statistical model, which is convenient for testing purposes, but does not require engagement at a level that would allow for testing the viewer's understanding of the information shown in the chart. 
<!-- An example of a statistical lineup for testing the perception of exponential growth is shown in @fig-lineup-log. -->

<!-- ![A lineup experiment testing whether users can perceive differences in rates of exponential growth on linear and log scales. In addition to asking users to select the plot which is the most different, we also asked for basic reasoning and the user's level of confidence in their answer.](Lineup_log_screenshot.png){#fig-lineup-log} -->

To assess the viewer's *understanding* of information shown in a chart, we must ask questions and allow the user to provide feedback. 
User feedback may be collected on a numerical scale or through the use of written comments, recorded "think-aloud" processes, and other more qualitative interaction methods. 
In some studies, asking users to interpret a chart within a larger scenario can be effective, as in @fig-estimation-describe, while in others it is more helpful to ask users to explain answers. 
In lineup studies, asking users why a specific panel was chosen can provide rich insight into otherwise confusing numerical results. 
While we have not (to date) recorded users talking out loud about what they are seeing, think aloud methods could be implemented within a Shiny application, with audio recordings saved to the server for transcription and analysis
[@dunbar1995scientists;@traftonTurningPicturesNumbers2000;@kirschenbaum2003comparative]; it is even possible that these recordings could be automatically transcribed using text-to-speech functions of large language models.

@@

Of course, in an online, asynchronous experiment, every user interaction with the testing materials (typically hosted on a web page) can also be recorded along with time stamps, mouse positions, browser size and screen resolution, and other information. 
While we have not used this type of information heavily in our experimental analyses thus far, in most experiments, we collect at least some time stamp data in order to assess how long participants spend on each question. 
Typically, the first round of test questions takes the longest for participants to complete. Additional replicates do not affect accuracy (i.e. there is no immediate learning effect) until after `too many' tests cognitive fatigue proves to be detrimental to accuracy [@chowdhury2018]. This sweet spot between replicates and fatigue depends on the cognitive burden in each test and should factor into designing the experiment.
<!--While we have not used this type of information heavily in our experimental analyses thus far, we do collect some time stamp data in order to assess how long participants spend on each question; it could almost certainly be more informative in the right setting.-->
In some experiments, we have provided participants with supportive tools, such as "scratch pads" and calculators built into the Shiny application to support the complex calculations  required to answer higher-level numerical estimation questions (@fig-estimation-calc). 
These supportive tools reduce participant cognitive load, but also provide a wealth of (messy and difficult to analyze) information support real insight into how participants were looking at the data, what strategies they tried and discarded for reading the chart, and what visual estimation methods were used.

@@

One of the most difficult components of designing an experiment which asks users to directly estimate information from a chart using a full scenario (background information, etc. as well as contextual details from the chart) is that the questions must be extremely carefully constructed.
Mathematics education researchers provide guidelines for selecting different levels of questioning in order to assess graph comprehension: literal reading of the data, reading between the data, and reading beyond the data [@wood1968objectives;@curcio1987comprehension;@friel2001making;@glazer2011challenges]. In a recent study, we identified questions based on this framework to evaluate direct estimates and extend those estimates to make comparisons between two points.

Even when great care is taken with the construction of the question, participant answer accuracy is fundamentally limited by the fact that many participants do not read and interpret the question with the care and precision that it was written.
Questions that ask participants to e.g. estimate the multiplicative change in a quantity at two time points may be misunderstood as asking for an estimate of the additive difference, and the resulting estimates are then one or more orders of magnitude off of the correct answer.
This is one area where lineup methods are convenient - they do not depend on participants to understand the nuances of language or scenarios built around the chart under investigation.
However, in some situations it may be sufficient to ask participants to estimate direct numerical quantities that have little contextual information, as we did in [@vanderplasFramedReproducingRevisiting2019] when assessing the accuracy of framed plots re-created from the Statistical Atlas.

It is also possible to design a task that requires participants to *engage directly* with an interactive visualization. 
This can be useful in a directed task, where users are asked to interact with the chart in a specific way and the result is recorded, but it is also possible to use interactive visualizations in an open-ended task, recording how users engage with the graphic in an exploratory (as opposed to goal-directed) manner. 
In one recent experiment, we asked participants to forecast an exponential trend, with data presented on either a linear or log scale. 
Using JavaScript code modified from New York Times interactive graphics "You Draw It" features [@katzYouDrawIt2017], we had users draw trend lines with their computer mouse and make forecasts directly on interactive charts, with the data and user-drawn predictions recorded to our database.
With interactive graphics rendered using JavaScript (or other web libraries), the only limit to the types of questions one can ask in testing graphics is one's ability to write code to interact with the visualization library. 
This type of testing method can be extremely natural for participants, but it also is hard to generalize when discussing testing methods because of the potential range of applications where it might be employed.

@@

Once the graphical task has been identified, it is necessary to develop a model which can be used to explore the graphical features of interest in a precise manner. 
This is the single longest part of the entire experimental design and execution process, in part because choosing a model that replicates important visual features of the data is  extremely complex [@hullmanDesigningInteractiveExploratory2021;@cookFoundationAvailableThinking2021;@vanderplasDesigningGraphicsRequires2021].

There are two main options when developing a statistical model for graphical testing: start with a large data set and sample from that data set [@power], or start from a model and sample data from that model generating process [@sineillusion;@clusters;@emily-diss].
This decision is largely determined by the availability of a large data set containing the requisite features of interest and the qualities being manipulated in the experiment. 
For instance, @power used samples of different sizes from a pre-existing data set to manipulate the amount of signal in each comparison; with a small sample, there is less signal and the same amount of noise, making the true plot harder to spot. 
In many situations, though, a convenient data set with the right properties is harder to acquire, and it becomes necessary to develop a sampling model to generate data for user evaluation.

The tools we discuss in the remainder of this section can be applied both to pre-existing data sets and to model-based sampling methods.

@@

Once an appropriate set of parameters are identified using the numerical screening method, then it is important to calibrate these parameter selections visually; after all, no numerical statistic is a perfect measure of what we actually see; at best, they are approximations of what we might potentially see. 
We have found it to be useful to have one experimenter calibrate the model parameters at a gross level, and then have another experimenter narrow in on the parameters which are visually reasonable within the selected range. Then, both examiners visually inspect a large number of plots generated using those parameters to get a sense for how difficult the task at hand is (this strategy is also described by Lu et al. [@luModelingJustNoticeable2022]). 
At some point, all experimenters become so visually saturated with the nuances of the data generating mechanism that it may become necessary to "sanity check" the protocol with family members, friends, and colleagues. 
These informal surveys provide extremely useful feedback and can help to counteract the visual saturation of being immersed in the design of a visualization experiment for months at a time.

@@

We cannot overstate the importance of visual assessment of your model stimuli, preferably with fresh eyes. We highly recommend performing a series of think-aloud pilot testing before deploying an experiment. In support of this assessment, we offer up a cautionary tale of our own experience: that of [@clusters], where we designed an experiment to test which plot aesthetics promoted discovery of linear trends and/or clusters.

The experiment was a full 2x3x3 factorial exploration of three data generating parameters, with 3 replicates at each parameter combination (54 data sets) and 10 aesthetic combinations (for a total of 540 lineups). Each lineup had 20 different sub-panels, so we should have carefully visually inspected some 10800 different panels. 
As is evident from the fact that we're telling this story as a cautionary tale, we missed a critical problem with our data-generating mechanism: when clusters were assigned to randomly generated data after the fact, we didn't control the cluster size, leading to clusters of one or two points in relatively few sub-panels. 
This became particularly noticeable when bounding ellipses were added to the plot, as the method used to generate those ellipses required at least 3 points in the cluster. The missing boundary ellipse in the corresponding sub-panels escaped our notice during the stimuli proof-reading phase of the experiment, but did not escape the notice of our participants, who only needed to examine about 10 lineups each (around 200 panels). 
An example of one of the problematic lineups is shown in @fig-lineup-problems: many participants selected panel 16 because of the missing ellipse; not a wrong choice, but certainly not the effect we intended to test.

![A lineup from Vanderplas & Hofmann [@clusters]. Panel 10 shows the clustered target data and panel 17 shows the target data with a strong linear relationship; either of these target panels was the expected choice. Unfortunately, panel 16 has only two bounding ellipses shown, which is an unintentional difference that resulted from a faulty method for assigning clusters to null plots; many participants selected this panel instead of one of the target panels.](lineup-missing-ellipse.png){#fig-lineup-problems}

@@

While studies have found some relationship between lineup performance and demographic factors [@majumderHumanFactorsInfluencing2014;@vanderplasSpatialReasoningData2016a], these differences are relatively small and unlikely to be important. 
Studies have also not found a strong relationship between visual task performance and participant recruitment method [@vanderplasFramedReproducingRevisiting2019], perhaps in part because demographics which use social media sites and demographics completing online research tasks for pay overlap heavily. 
What is clear, however, is that some participants are very good at evaluating graphics and some participants are not; as a result, it can be extremely helpful to use random effects models with an effect for individual participants.

@@

**Generalized Linear Mixed Models** 
 The see-value approach [@niladriroychowdhurySeeValueApp2020;@vanderplasStatisticalSignificanceCalculations2021;@majumderValidationVisualStatistical2013] is extremely useful for single lineups, but when a series of lineups that are part of a designed experiment are used, generalized linear mixed models are a much simpler way to summarize the overall effect of various manipulated factors [@power;@clusters;@emily-diss]. This approach also works extremely well for psychometric experiments [@sineillusion], as psychometric models can be easily fit into the framework of a generalized linear mixed model. In addition, similar model structures can be used to model accuracy, response time, and confidence, if all three types of information are collected from participants during the experiment.

**Numerical Estimation** There are additional considerations that should be expected when asking participants to estimate numerical quantities. Anchoring and rounding cause participant responses to cluster in ways that can bias statistical estimators, requiring methods designed for these types of data [@ushakovStatisticalAnalysisRounded2017;@tourangeau_rips_rasinski_2000;@heitjanIgnorabilityCoarseData1991]. An alternative approach is to analyze the data graphically using e.g. a combination of density and rug geoms for point estimates (as shown in @fig-density-rug), or spaghetti plots for individual response curves, with annotations to show anchor points. The advantage of this approach is that it can accommodate extremely messy data without requiring the extensive data cleaning and elimination of nonsensical responses that might be necessary to fit a statistical model.

@@

grid_lines_data <- tibble(scale = c(rep("linear", 12), rep("log2", 10)), 
                          grid_lines = c(seq(0,55000, 5000),
                                         2^seq(7,16))
) %>%
  expand_grid(data set = c("data set1", "data set2"))

@@

```{r density-plot, include = F, echo = F}
densityPlot <- function(data, data setID, estimate, xlabel = "Estimate", x_limits = c(0,70000), zoom = F, scalesx = T, zoom_limits = c(NA,NA), gridlines = T, rugjitter = 0.25) {

  estPlot <- data %>%
    filter(data set %in% data setID) %>%
    ggplot(aes_string(x = estimate, fill = "scale", color = "scale")) +
    geom_density(alpha = 0.5, color = NA) +
    geom_rug(aes(y = -Inf), alpha = 0.6, show.legend = F, position = position_jitter(width = rugjitter, height = 0)) +
    geom_vline(aes(xintercept = true_value, linetype = "a")) +
    geom_vline(aes(xintercept = closest_pt_value, linetype = "b"))

  if (gridlines) {
  estPlot <- estPlot + 
    geom_vline(data = grid_lines_data %>% filter(data set == "data set1", grid_lines >= x_limits[1], grid_lines <= x_limits[2]),
               aes(xintercept = grid_lines, color = scale, linetype = "c"))
  }
